Algebraic Tori in Unitary Groups

### 2014-03-23 Paper.tex
add some more background

\TODO{} Decide which of the following to define/give notations for, we probably don't need all of it. Some of it can be postponed until the sections where it is used.
\begin{itemize}
\item Explain our notation for fields and \'etale algebras: $\aK$, $\gK$, $\lK$, $\fK$, $\aL$, $\gL$, $\lL$, $\fL$, $\aE$, $\gE$, $\lE$, $\fE$, $\aV$, $\gV$, $\lV$, $\fV$.
\item The setting is a base field $K$ taken to be a $p$-adic field or number field.
\item We shall denote by $\GalKbK$ the absolute Galois group of $K$.
\item Algebraic Tori are... 
\item \'Etale algebras are... typically denoted by $E$.  Write $T_E$ for the $K$-torus $\Res_{E/K} \Gm$.
\item \'Etale algebras with involutions are denoted by $(E, \sigma)$.  Write $T_{E, \sigma}$ for the $K$-torus defined as the restriction from $E^\sigma$ to $K$ of the Norm 1 torus in $\Res_{E/E^\sigma} \Gm$.
\item Hermitian Spaces are ... their invariants over (local fields, global fields) are...
\item Unitary groups are ... their invariants over (local fields, global fields) are...
\item Quadratic Forms are ...their invariants over (local fields, global fields) are...
\item Orthogonal Groups are ...their invariants over (local fields, global fields) are...
\item Connections between orthogonal and unitary groups.
\end{itemize}

>>>>

\TODO{} Decide which of the following to define/give notations for, we probably don't need all of it. Some of it can be postponed until the sections where it is used.
\begin{itemize}
\item Explain our notation for fields and \'etale algebras: $\aK$, $\gK$, $\lK$, $\fK$, $\aL$, $\gL$, $\lL$, $\fL$, $\aE$, $\gE$, $\lE$, $\fE$, $\aV$, $\gV$, $\lV$, $\fV$.
\item The setting is a base field $K$ taken to be a $p$-adic field or number field.
\item We shall denote by $\GalKbK$ the absolute Galois group of $K$.
\item Algebraic Tori are... 
\item \'Etale algebras are... typically denoted by $E$.  Write $T_E$ for the $K$-torus $\Res_{E/K} \Gm$.
\item \'Etale algebras with involutions are denoted by $(E, \sigma)$.  Write $T_{E, \sigma}$ for the $K$-torus defined as the restriction from $E^\sigma$ to $K$ of the Norm 1 torus in $\Res_{E/E^\sigma} \Gm$.
\item

Fix a quadratic \'etale extension $E$ of $K$ and a free $E$-module $V$.
A Hermitian structure on $V$ is a non-degenerate skew form $H : V\times V \rightarrow E$.
We shall consider the group $U(H) = \Aut(V,H)$ and $SU(H) = \Aut(V,H,\det_E)$ where $\det_E$ refers to the $E$-module determinant on $V$.

The forms of $V$ are in bijection with $H^1(\GalKbK, \Aut(V,H))$.
We have an exact sequence:
\[ 1 \rightarrow SU(H) \rightarrow U(H) \rightarrow T_{E,\sigma} \rightarrow 1. \]
This induces an exact sequence:
\[ \ldots \rightarrow H^1(\GalKbK,SU(H))  \rightarrow H^1(\GalKbK,U(H))  \rightarrow H^1(\GalKbK,T_{E,\sigma}) \rightarrow \ldots. \]
We have an isomorphism $H^1(\GalKbK,T_{E,\sigma}) = K^\times/N_{E/K}(E^\times)$ and we shall refer to the associated invariant of $H$ as the discriminant.
The group $SU(H)$ is simply connected, thus its cohomology is supported at the real places of $K$.
When $E$ is split there are no forms, when $E=\C$ forms are classified by their signatures, that is the number of positive and negative eigenvalues for their Grahm matrix.

We summarize the above with the following well known proposition.
\begin{proposition}
Fix a quadratic \'etale extension $E$ of $K$ and a free $E$-module $V$.
Hermitian structures on $E$ are classified up to isomorphism by their discriminants $\delta\in K^\times/N_{E/K}(E^\times)$ and their signatures $(s_\nu,r_\nu)$ at real places of $K$ which ramify in $E$.
Moreover, $\delta$ may be computed as the determinant of the Grahm matrix whereas $s_\nu$ and $r_\nu$ are respectively the number of positive and negative eigenvalues of hte Grahm matrix.
\end{proposition}

\item

For our purposes a unitary group is a simply connect group $G$ of type $A_n$.

The forms of $G$ are in bijection with the elements of $H^1(\GalKbK, \Aut(G))$.

When $n>1$ the outer automorphism is a group of order $2$ and we have an exact sequence:
\[ 1 \rightarrow   G^{adj} \rightarrow \Aut(G) \rightarrow \mu_2  \rightarrow 1 \]
from which we obtain an exact sequence:
\[ \ldots  H^1(\GalKbK, G^{adj})  \rightarrow  H^1(\GalKbK, \Aut(G)) \rightarrow  H^1(\GalKbK, \mu_2) \rightarrow \ldots \]
We conclude we can characterize forms of $G$ by invariants in $H^1(\GalKbK, \mu_2)$ and $H^1(\GalKbK, G^{adj})$.

An element of $H^1(\GalKbK, \mu_2)$ corresponds to a quadratic \'etale extension $E$ of $K$.
More precisely it is the $n+1$ torsion of $T_{E,\sigma}$.
We have an exact sequence:
\[ \ldots \rightarrow H^1(\GalKbK,G) \rightarrow H^1(\GalKbK, G^{adj}) \rightarrow H^2(\GalKbK, \mu_{n+1}^\xi) \rightarrow \ldots. \]
A form of $G$ thus has an invariant in $H^2(\GalKbK, mu_{n+1}^\xi)$.

$H^2(\GalKbK, \mu_{n+1}^\xi)$ fits into the exact sequence:
\[ H^1(\GalKbK, T_{E,\sigma}) \rightarrow H^2(\GalKbK, mu_{n+1}^\xi) \rightarrow H^2(\GalKbK, T_{E,\sigma}) \overset{[n+1]}\rightarrow H^2(\GalKbK, T_{E,\sigma}). \]
We are thus interested in the cohomology of $T_{E,\sigma}$.

The exact sequence $1 \rightarrow T_{E,\sigma} \rightarrow T_E \rightarrow T_{E^\sigma} \rightarrow 1$ and Hilbert's Theorem 90 allow us to compute that 
$H^1(\GalKbK, T_{E,\sigma}) = k^\times/N_{E/k}(E^\times)$ and that $H^2(\GalKbK, T_{E,\sigma})$ is the kernel of the corestriction map from the Brauer group of $E$ to that of $E^\sigma$.
In particular we can now deduce that $H^2(\GalKbK, \mu_{n+1}^\xi)$ corresponds to element a pair of an element of $\delta\in k^\times/(k^\times)^nN_{E/k}(E^\times)$ and a degree $n$ central simple algebra over $E$ in the $n$-torsion of the Brauer group which is only non-trivial at the places of $E$ over $K$ which split, and for places which split the algebra over one factor is isomorphic to the opposite algebra of that over the other.

\begin{remark}
It should be pointed out that the group $K^\times/(K^\times)^nN_{E/K}(E^\times)$ is trivial if either $n$ is odd or if $E = K\times K$ is split.
\end{remark}

Finally, we must consider the group $H^1(\GalKbK,G)$. As $G$ is simply connect, this is only non-trivial at the real places of $K$.
In terms of real places, there are two cases to consider, when $E=\RR\times\RR$ then the $G$ is the automorphism group of a free module.
 As there are no forms of free modules, $H^1(\GalKbK,G)$ is then trivial. If $E=\C$, then $G$ is the automorphism group of a Hermitian space and $H^1(\GalKbK,G)$ classifies the Hermitian forms of the same discriminant as that defining $G$, over $\RR$ this means it classifies the signature.

Finally, we must consider the group $H^1(\GalKbK,G)$. As $G$ is simply connect, this is only non-trivial at the real places of $K$.
More concretely, as $\Gal(\C/\RR)$ has a single element, $H^1(\GalKbK,G)$ is determined by $g\in G$ such that $g\sigma(g) = 1$ modulo $hg\sigma(h^{-1})$.
TODO ---- do we actually want to explain this part concretely

\begin{remark}
Because we are dealing with non-abelian Galois cohomology when analyzing exact sequences we must twist by the cokernel elements before considering the structure of the kernel. It is for this reason that we must look at the invariants in the natural order $E$, $A$, $\delta$, signatures.
\end{remark}

Given a a collection of invariants, $E$, $A$, $\delta$, and $(s_\nu,r_\nu)$ we can construct the group $G$ as follows.
By assumption $\sigma(A) \simeq A^{op}$ and thus $A$ can be given an involution $\sigma_A$ restricting to $\sigma$ on $E$.
Fix an element $J \in A^\sigma$ of reduced norm $\delta$ such that at each real place of $K$ which ramifies in $E$ the matrix for $J$ has $r$ positive eigenvalues and $s$ negative eigenvalues where $(r,s)$ is the signature at the real place. We may then define $G$ as:
\[ G(R) = \{ g \in (A\otimes_K R)^\times \mid gJ\sigma_A(g) = J \} \]
One may check that $G$ determines $E$, $A$, $\delta$ and the signatures uniquely when $n>1$.

\begin{remark}
In the case of $n=1$, we may perform the same constructions, however the group does not uniquely determine invariants $E$, $A$, $\delta$, and signatures.
Indeed, there is no outer automorphism group, so the choice of $E$ is no longer an invariant.
Exceptional isomorphisms of groups tell us that for $n=1$ the group $G$ is the spin group of a rank $3$ quadratic space.
In odd rank the discriminant does not impact the isomorphism type of the orthogonal group.
The discriminant of the quadratic space does still play a role in defining the Clifford algebra, in the same way it plays a role in defining a unitary group. The Witt invariant of the quadratic space is still an invariant and determines the algebra $A$.

It is worth remembering that choosing an $E$ and a discriminant is equivalent to choosing $A$.

The following example illustrates that any two quadratic algebras have $2$-dimensional Hermitian spaces with isomorphic special unitary groups.
\begin{example}
Consider $L = K(\sqrt{d_1},\sqrt{d_2})$ with $\sigma$ the involution which acts as $-1$ on both $\sqrt{d_1}$ and $\sqrt{d_2}$.
Then the maps $H_1 : L \times L \rightarrow K(\sqrt{d_1})$  $H_2 : L \times L \rightarrow K(\sqrt{d_2})$ given respectively by 
$H_i(x,y) =  Tr_{L/K(\sqrt{d_i})}(\sqrt{\delta_1\delta_2}x\sigma(y))$ gives $L$ the structure of $K(\sqrt{d_i})$-Hermitian spaces.

We may also consider the $K$-quadratic structure $B(x,y) = Tr_{L/K}(\sqrt{\delta_1\delta_2}x\sigma(y))$ and Symplectic structures $S_i(x,y) = Tr_{L/K}(\sqrt{\delta_i}x\sigma(y))$.
Then we observe that $\Aut(H_i) = \Aut(B) \cap \Aut(S_i)$. With respect to the basis $\{1,\sqrt{\delta_1},\sqrt{\delta_2},\sqrt{\delta_1\delta_2}\}$ the forms have Gram matricies:
\[ B \sim \begin{pmatrix} 0 & 0 & 0 & \delta_1\delta_2 \\ 0 & 0 &  \delta_1\delta_2 & 0 \\ 0 &  \delta_1\delta_2 & 0 & 0 \\  \delta_1\delta_2 & 0 & 0 & 0 \end{pmatrix} \]
\[ S_1 \sim \begin{pmatrix} 0 & \delta_1 & 0 & 0 \\ -\delta_1 & 0 & 0 & 0 \\ 0 & 0 & 0 & \delta_1\delta_2 \\   0 & 0 & -\delta_1\delta_2 & 0  \end{pmatrix} \]
\[ S_2  \sim \begin{pmatrix} 0 & 0 &  \delta_1&0 \\ 0 & 0 & 0 & \delta_1\delta_2 &  \\  -\delta_1 & 0  & 0 & 0 \\   0 & -\delta_1\delta_2 &0 & 0  \end{pmatrix} \]
The matrix $\begin{pmatrix} 1 & 0 & 0 & 0 \\ 0 & 0 & 1 & 0 \\ 0 & 1 & 0 & 0 \\  0 & 0 & 0 & 1 \end{pmatrix} \in \Aut(B)$ transforms $S_1$ to $S_2$ and thus transforms $\Aut(H_1)$ to $\Aut(H_2)$
\end{example}
\end{remark}

\item Quadratic Forms are ...their invariants over (local fields, global fields) are...
\item Orthogonal Groups are ...their invariants over (local fields, global fields) are...
\item Connections between orthogonal and unitary groups.
\end{itemize}
